---
title: "Week 6 Double Machine Learning for Partially Linear IV Model"
format:
  pdf:
    pdf-engine: pdflatex 
    documentclass: article
    classoption: ["12pt"]
    geometry: "top=1.1in, bottom=1.1in, left=1.1in, right=1.1in"
    include-in-header:
      text: |
        \DeclareMathOperator*{\argmin}{arg\,min}
        % fonts, math
        \usepackage[T1]{fontenc}
        \usepackage{dsfont}
        \usepackage{newpxtext}
        \let\arrowvert\relax
        \let\Arrowvert\relax
        \let\mathscr\relax
        \usepackage{newpxmath}
        \usepackage{amsmath}
        \usepackage{bm}
        
        % colors
        \usepackage[dvipsnames]{xcolor}
        
        % Indent & line spacing
        \usepackage{setspace}
        \setstretch{1.2}
        \setlength{\parindent}{20pt}
        \AtBeginDocument{
        \setlength{\abovedisplayshortskip}{-10pt}
        \setlength{\belowdisplayshortskip}{3.5pt}
        }
        
        % Figures
        \usepackage{graphicx}
        \usepackage{tikz}
        \usetikzlibrary{arrows.meta, positioning, fit}

        % Equal spacing between Title, Date, Author
        \makeatletter
        \renewcommand{\maketitle}{%
          \begin{center}
            {\LARGE \bfseries \@title \par}
            \vspace{0.75em}%
            {\large \itshape \@date \par}
            \vspace{0.75em}%
            {\large \@author \par}
          \end{center}\par\vspace{1em}
        }
        \makeatother
execute:
  echo: true
  warning: false
  message: false
editor: source
---

# Partially Linear IV Model

We consider the partially linear structural equation model

$$
\begin{aligned}
 Y_i :=~& \theta_0 D_i + g_0(X_i) + \zeta_i,  & E[\zeta_i \mid Z_i,X_i]= 0 \\
 Z_i :=~& m_0(X_i) +  \epsilon_i,   &  E[\epsilon_i \mid X_i] = 0
\end{aligned}
$$

where there is a structural or causal relation between $D_i$ and $Y_i$, captured by $\theta_0$. $g_0(X_i) + \zeta_i$ is the stochastic error, partly explained by covariates $X_i$.  $\epsilon_i$ and $\zeta_i$ are stochastic errors that are not explained by $X_i$. $Y_i$ and $D_i$ are jointly determined, meaning that $D_i$ is *endogeneous*, with $E[\zeta_i \mid D_i] \neq 0$. The instrument, $Z_i$, is introduced to create exogenous variation
in $D_i$. The observed $X_i$ serve as confounding factors, and we can think of variation in $Z_i$ as being exogenous conditional on $X_i$.


# PLIV Model in Residualized Form

The PLIV model above can be rewritten in the following residualized form

$$
  \tilde Y_i = \theta_0 \tilde D_i + \zeta_i,   \quad  E[\zeta_i \mid \epsilon_i,X_i]= 0
$$
where
$$
\begin{aligned}
 \tilde Y_i &= (Y_i - \ell_0(X_i)),  \quad &\ell_0(X_i) &= E[Y_i \mid X_i] \\
   \tilde D_i &= (D_i - r_0(X_i)), \quad &r_0(X_i) &= E[D_i \mid X_i] \\
   \tilde Z_i &= (Z_i - m_0(X_i)), \quad &m_0(X_i) &= E[Z_i \mid X_i]
\end{aligned}
$$

# DML for PLIV Model

Given the valid instrument for identification, DML proceeds as follows.

Compute the estimates $\hat \ell_0$, $\hat r_0$, and $\hat m_0$ , which amounts
to solving the three problems of predicting $Y_i$, $D_i$, and $Z_i$ using
$X_i$, using any generic ML methods. Then estimate $\theta_0$ by the the instrumental variable regression of $\tilde Y$ on $\tilde D$ using $\tilde Z$ as an instrument.
Use the conventional inference for the IV regression estimator, ignoring
the estimation error in these residuals.

```{r load package, warning=FALSE, message=FALSE}
library(hdm)
library(AER)
library(randomForest)
library(lfe)
library(glmnet)
set.seed(1)
```

```{r DML for PLIV Model, warning=FALSE, message=FALSE}

## DML for PLIV Model
dml2_for_plivm <- function(x, d, z, y, dreg, yreg, zreg, nfold = 5) {
  
  ## cross-fitting
  nobs <- nrow(x)
  foldid <- rep.int(1:nfold, times = ceiling(nobs / nfold))[sample.int(nobs)]
  I <- split(1:nobs, foldid)
  
  ## create residualized objects to fill
  ytil <- dtil <- ztil <- rep(NA, nobs)
  
  ## obtain cross-fitted residuals
  cat("fold: ")
  for (b in seq_along(I)) {
    dfit <- dreg(x[-I[[b]], ], d[-I[[b]]]) ## take a fold out
    zfit <- zreg(x[-I[[b]], ], z[-I[[b]]]) ## take a fold out
    yfit <- yreg(x[-I[[b]], ], y[-I[[b]]]) ## take a fold out
    dhat <- predict(dfit, x[I[[b]], ], type = "response") ## predict the held-out fold
    zhat <- predict(zfit, x[I[[b]], ], type = "response") ## predict the held-out fold
    yhat <- predict(yfit, x[I[[b]], ], type = "response") ## predict the held-out fold
    dtil[I[[b]]] <- (d[I[[b]]] - dhat) # record residual
    ztil[I[[b]]] <- (z[I[[b]]] - zhat) # record residual
    ytil[I[[b]]] <- (y[I[[b]]] - yhat) # record residual
    cat(b, " ")
  }
  ivfit <- tsls(y = ytil, d = dtil, x = NULL, z = ztil, intercept = FALSE)
  coef_est <- ivfit$coef ## extract coefficient
  se <- ivfit$se ## record standard error
  cat(sprintf("\ncoef (se) = %g (%g)\n", coef_est, se))
  return(list(coef_est = coef_est, se = se, dtil = dtil, ytil = ytil, ztil = ztil))
}
```


# Emprical Example:  Acemoglu, Johnson, Robinson (AER)

* $Y_i$ is log modern GDP;
* $D_i$ is a measure of protection from expropriation, a proxy for quality of institutions;
* $Z_i$ is the log of settler's mortality;
* $W_i$ are geographical variables (latitude, latitude squared, continent dummies as well as interactions)

```{r read AJR, warning=FALSE, message=FALSE}
## read AJR data
file <- "https://raw.githubusercontent.com/CausalAIBook/MetricsMLNotebooks/main/data/AJR.csv"
AJR <- read.csv(file)
dim(AJR)

## construct variables
y <- AJR$GDP
d <- AJR$Exprop
z <- AJR$logMort
xraw <- model.matrix(~ Latitude + Africa + Asia + Namer + Samer, data = AJR)
x <- model.matrix(~ -1 + (Latitude + Latitude2 + Africa + Asia + Namer + Samer)^2, data = AJR)
dim(x)
```

```{r standard IV estimation, message=FALSE, warning=FALSE}

## IV specification and weak IV test
library(ivreg)
fit <- ivreg(y ~ d + Latitude + Africa + Asia + Namer + Samer | 
               z + Latitude + Africa + Asia + Namer + Samer, 
             data = AJR)
summary(fit, diagnostics = TRUE)
```


```{r application of DML for PLIV Model, message=FALSE, warning=FALSE}

set.seed(1)

## DML with Post-Lasso
cat(sprintf("\n DML with Post-Lasso \n"))

dreg <- function(x, d) {
  hdm::rlasso(x, d, post=TRUE) ## default plug-in penalty level
} 
yreg <- function(x, y) {
  hdm::rlasso(x, y, post=TRUE) ## default plug-in penalty level
}
zreg <- function(x, z) {
  hdm::rlasso(x, z, post=TRUE) ## default plug-in penalty level
}

## specify the function
dml2_lasso <- dml2_for_plivm(x, d, z, y, dreg, yreg, zreg, nfold = 5)

# DML with Random Forest
cat(sprintf("\n DML with Random Forest \n"))

dreg <- function(x, d) {
  randomForest(x, d) ## default ntrees = 500
}
yreg <- function(x, y) {
  randomForest(x, y) ## default ntrees = 500
}
zreg <- function(x, z) {
  randomForest(x, z) ## default ntrees = 500
}

## specify the function
dml2_rf <- dml2_for_plivm(xraw, d, z, y, dreg, yreg, zreg, nfold = 5)

## compare Forest vs Lasso
comp_tab <- matrix(NA, 3, 2)
comp_tab[1, ] <- c(sqrt(mean((dml2_rf$ytil)^2)), sqrt(mean((dml2_lasso$ytil)^2)))
comp_tab[2, ] <- c(sqrt(mean((dml2_rf$dtil)^2)), sqrt(mean((dml2_lasso$dtil)^2)))
comp_tab[3, ] <- c(sqrt(mean((dml2_rf$ztil)^2)), sqrt(mean((dml2_lasso$ztil)^2)))
rownames(comp_tab) <- c("RMSE for Y:", "RMSE for D:", "RMSE for Z:")
colnames(comp_tab) <- c("RF", "LASSO")
print(comp_tab, digits = 3)
```

# Weak Instrument Problem

```{r weak instrument, warning=FALSE, message=FALSE}

## using lfe package
summary(felm(dml2_lasso$dtil ~ dml2_lasso$ztil), robust = TRUE)
summary(felm(dml2_rf$dtil ~ dml2_rf$ztil), robust = TRUE)
```

## Anderson-Rubin Idea for Inference with Weak Instruments

As shown above, we may have weak instruments because the F-statistics in the regression $\tilde D_i \sim \tilde Z_i$ is small relative to standard rules of thumb -- for example Stock and Yogo (2005) suggest accounting for weak instruments if the first stage F-statistic is less than 10 (and more recent work suggests even larger cutoffs).


Here, we consider one specific approach (from Anderson and Rubin (1949)) for valid inference under weak identification based upon the statistic:

$$
C(\theta) = \frac{ |\mathbb{E}_n [(\tilde Y -  \theta  \tilde D) \tilde Z]|^2}{ \mathbb{V}_n [(\tilde Y -  \theta \tilde  D) \tilde Z ]/n}
$$
The empirical variance $\mathbb{V}_n$ is defined as

$$
\begin{aligned}
\mathbb{V}_{n}[ g(W_i)] &:=   \mathbb{E}_{n}g(W_i)g(W_i)' - \mathbb{E}_{n}[ g(W_i)]\mathbb{E}_{n}[ g(W_i)]'
\end{aligned}
$$

If $\theta_0 = \theta$, then $C(\theta) \xrightarrow{d} \mathcal{N}(0,1)^2 = \chi^2(1)$. Therefore,  we can reject the hypothesis $\theta_0 = \theta$ at level $a$ (for example $a = .05$ for a 5\% level test)  if $C(\theta)> c(1-a)$ where $c(1-a)$ is the $(1-a)$- quantile of a $\chi^2(1)$ variable. The probability of falsely rejecting the true hypothesis is approximately $a \times 100\%$.  

To construct a $(1-a) \times 100$\% confidence region for $\theta$, we can then simply invert the test by collecting all parameter values that are not rejected at the $a$ level:
$$
CR(\theta) = \{ \theta \in \Theta: C(\theta)  \leq c(1-a)\}
$$

```{r implementation of DML-AR, warning=FALSE, message=FALSE}

## DML-AR (DML with Anderson-Rubin)
dml_ar_pliv <- function(rY, rD, rZ, grid, alpha = .05) {
  n <- length(rY)
  Cstat <- rep(0, length(grid))
  for (i in seq_along(grid)) {
    Cstat[i] <- n * (mean((rY - grid[i] * rD) * rZ))^2 / var((rY - grid[i] * rD) * rZ)
  }
  LB <- min(grid[Cstat < qchisq(1 - alpha, 1)])
  UB <- max(grid[Cstat < qchisq(1 - alpha, 1)])
  plot(range(grid), range(c(Cstat)), type = "n", xlab = "Effect of institutions", ylab = "Statistic", main = " ")
  lines(grid, Cstat, lty = 1, col = 1)
  abline(h = qchisq(1 - alpha, 1), lty = 3, col = 4)
  abline(v = LB, lty = 3, col = 2)
  abline(v = UB, lty = 3, col = 2)
  return(list(UB = UB, LB = LB))
}

dml_ar_pliv(
  rY = dml2_lasso$ytil, rD = dml2_lasso$dtil, rZ = dml2_lasso$ztil,
  grid = seq(-2, 2, by = .01)
)
dml_ar_pliv(
  rY = dml2_rf$ytil, rD = dml2_rf$dtil, rZ = dml2_rf$ztil,
  grid = seq(-2, 2, by = .01)
)
```

For any $\theta$ with $C(\theta) \leq 3.84$, we cannot reject the null hypothesis that $\theta = \theta_0$ at the 5% level. Those $\theta$ values form the Anderson–Rubin confidence set.
