---
title: "\\textbf{Causal Machine Learning: Homework 1}"
subtitle: "\\textit{Due date}: September 22, 11:59PM"
author: "Your name:"
format:
  pdf:
    pdf-engine: pdflatex 
    documentclass: article
    classoption: ["12pt"]
    geometry: "top=1.1in, bottom=1.1in, left=1.1in, right=1.1in"
    include-in-header:
      text: |
        \DeclareMathOperator*{\argmin}{arg\,min}
        % fonts, math
        \usepackage[T1]{fontenc}
        \usepackage{lmodern}
        \usepackage{amsmath}
        \usepackage{bm}
        
        % colors
        \usepackage[dvipsnames]{xcolor}
        
        % Indent & line spacing
        \usepackage{setspace}
        \setstretch{1.2}
        \setlength{\parindent}{20pt}
        \AtBeginDocument{
        \setlength{\abovedisplayshortskip}{-10pt}
        \setlength{\belowdisplayshortskip}{3.5pt}
        }

        % Equal spacing between Title, Date, Author
        \makeatletter
        \renewcommand{\maketitle}{%
          \begin{center}
            {\LARGE \bfseries \@title \par}
            \vspace{0.75em}%
            {\large \itshape \@date \par}
            \vspace{0.75em}%
            {\large \@author \par}
          \end{center}\par\vspace{1em}
        }
        \makeatother
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

# Bias-Variance Tradeoff

1.  In your own words, explain the bias–variance tradeoff when using a *sample* to approximate the *population* Conditional Expectation Function (CEF), and clarify where the variance in prediction comes from.

2.  In your own words, explain how *Random Forests* address the bias–variance tradeoff, and why balancing this tradeoff is important.

3.  In your own words, explain how *Boosted Trees* differ from *Bagging* and *Random Forests* in how they address the bias–variance tradeoff.

# Frisch–Waugh–Lovell Theorem

In lecture, we showed that the multivariate OLS regression of $Y_i$ against the treatment variable $D_i$ and control covariates $W_i$

$$
\begin{aligned}
Y_i &= \beta_1D_i + \beta_2'W_i + \epsilon_i
\end{aligned}
$$

\noindent can be recovered in a bivariate regression,

$$
\begin{aligned}
\tilde{Y}_i &= \beta_1\tilde{D}_i + \mu_i \\
\beta_1 &= \frac{Cov(\tilde{Y}_i,\tilde{D}_i)}{Var(\tilde{D}_i)}
\end{aligned}
$$

\noindent where $\tilde{D}_i$ and $\tilde{Y}_i$ are defined as residuals from their respective regressions against $W_i$.

$$
\begin{aligned}
\tilde{D}_i = D_i - \gamma_{DW}' W_i &, \quad
\tilde{Y}_i = Y_i - \gamma_{YW}' W_i \\
\gamma_{DW} = \argmin_{\gamma} E\left[ (D_i - \gamma'W_i)^2 \right] &, \quad
\gamma_{YW} = \argmin_{\gamma} E\left[ (Y_i - \gamma'W_i)^2 \right]
\end{aligned}
$$

1.  Show that $\beta_1$ can be similarly recovered by using $Y_i$ rather than residualized $\tilde Y_i$

$$
\beta_1 = \frac{Cov(Y_i,\tilde{D}_i)}{Var(\tilde{D}_i)}
$$

2.  Symmetrically, can one use unresidualized $D_i$ in the numerator to express $\beta_1$?

$$
\beta_1 \stackrel{?}{=} \frac{Cov(\tilde Y_i,D_i)}{Var(\tilde{D}_i)}
$$

3.  In a high-dimensional setting where $p/n \not\to 0$, the OLS estimator of $\beta_1$ is no longer well-behaved, and its sampling variability can be severely inflated. We introduced Double LASSO in lecture to address this high-dimensional problem.

<!-- -->

(a) Explain in words the relationship between the *penalty level* $\lambda$ in Double LASSO and the theoretical variance of your estimated $\beta_1$.

(b) After running Double LASSO, we obtain residualized $D_i$ and $Y_i$ by partialling out high-dimensional controls. In the final bivariate regression of residualized $Y_i$ on residualized $D_i$, can we rely on the standard sandwich formula for the standard error of $\hat\beta_1$? Why or why not?

# Test of Convergence Hypothesis

In lecture, we provided an empirical example of partialling-out with Double LASSO to estimate the regression coefficient $\beta_1$ in the high-dimensional linear regression model. Specifically, we estimated how the rates at which economies of different countries grow ($Y_i$) are related to the initial wealth levels in each country ($D_i$) controlling for country's institutional, educational, and other similar characteristics ($W$). The relationship is captured by $\beta_1$, the *speed of convergence/divergence*, which measures the speed at which poor countries catch up $(\beta_1<0)$ or fall behind $(\beta_1>0)$ rich countries, after controlling for $W_i$.

Indeed, residualized $D_i$ and $Y_i$ can also be estimated using Machine Learning (ML) methods beyond LASSO, as demonstrated in lecture. In this question, you will estimate residualized $D_i$ and $Y_i$ using the ML methods introduced in class. For this exercise, we will use the dataset `GrowthData`, which is included in the `hdm` package.  

```{r load data, warning=FALSE}
library(hdm)
## function to get data
getdata <- function(...) {
  e <- new.env()
  name <- data(..., envir = e)[1]
  e[[name]]
}

## now load your data calling getdata()
growth <- getdata(GrowthData)
```

```{r create variables, warning=FALSE}
## create the outcome variable y, treatment d, and control covariates W
y <- growth$Outcome
W <- growth[-which(colnames(growth) %in% c("Outcome", "intercept", "gdpsh465"))]
D <- growth$gdpsh465
```

1.  Using the package `randomForest` and its default settings of `ntree` and `nodesize`, report the estimated $\beta_1$. Use `set.seed(1)`. What can you conclude from your estimation?

2.  Now change the setting and let `ntree = 10` and `nodesize = 1`. Use `set.seed(1)`. Re-estimate $\beta_1$ and report your findings. In your own words, explain the possible reasons of changes in your findings.

```{r double random forest, warning=FALSE}
set.seed(1)
library(randomForest)

```

3.  We may also use a Neural Network model to estimate the residualized $D_i$ and $Y_i$. Set the random seed with `set.seed(1)`. Configure the hyperparameters as follows:

-   Build a neural network with one hidden layer containing 200 neurons.

-   Train the network using Stochastic Gradient Descent (SGD) with a `learning rate` of 0.05, 200 `epochs`, and a `batch size` of 10.

-   Implement early stopping based on performance on a 20% validation set, with a patience of 10 `epochs`.

\noindent Based on your estimation results $\beta_1$ under this model, what can you conclude about the relationship between country's initial wealth and growth rate?

```{r double neural network, warning=FALSE}
set.seed(1)
library(keras3)

```
