---
title: "Controls for IV Estimation"
format:
  pdf:
    pdf-engine: pdflatex 
    documentclass: article
    classoption: ["12pt"]
    geometry: "top=1.1in, bottom=1.1in, left=1.1in, right=1.1in"
    include-in-header:
      text: |
        \DeclareMathOperator*{\argmin}{arg\,min}
        % fonts, math
        \usepackage[T1]{fontenc}
        \usepackage{dsfont}
        \usepackage{newpxtext}
        \let\arrowvert\relax
        \let\Arrowvert\relax
        \let\mathscr\relax
        \usepackage{newpxmath}
        \usepackage{amsmath}
        \usepackage{bm}
        
        % colors
        \usepackage[dvipsnames]{xcolor}
        
        % Indent & line spacing
        \usepackage{setspace}
        \setstretch{1.2}
        \setlength{\parindent}{20pt}
        \AtBeginDocument{
        \setlength{\abovedisplayshortskip}{-10pt}
        \setlength{\belowdisplayshortskip}{3.5pt}
        }
        
        % Figures
        \usepackage{graphicx}
        \usepackage{tikz}
        \usetikzlibrary{arrows.meta, positioning, fit}

        % Equal spacing between Title, Date, Author
        \makeatletter
        \renewcommand{\maketitle}{%
          \begin{center}
            {\LARGE \bfseries \@title \par}
            \vspace{0.75em}%
            {\large \itshape \@date \par}
            \vspace{0.75em}%
            {\large \@author \par}
          \end{center}\par\vspace{1em}
        }
        \makeatother
execute:
  echo: true
  warning: false
  message: false
editor: source
---

## Validity of Instrumental Variable in the Special Case

In Assignment 3, we have a special scenario where there is an unobserved $V$ that affects $D$ and $Z$ (left panel). Can $Z$ be a valid instrument? Here I simulate the DGP and check the property of the estimate. Note the true $\beta$ in the DGP is 5.

\begin{center}
\begin{tikzpicture}[
>=stealth,
scale=0.65,
thick,
every node/.style={scale=1.1},
]

% nodes
\node (D) at (8,0) {$D$};
\node[circle,draw,dashed,inner sep=2pt] (U) at (9.5,-2.5) {$U$};
\node (Y) at (11,0) {$Y$};
\node[circle,draw,dashed,inner sep=2pt] (V) at (5,0) {$V$};
\node (Z) at (6.5,2.5) {$Z$};
\node (X) at (9.5,2.5) {$X$};

% arrows
\draw[->] (Z) -- (D);
\draw[->] (D) -- (Y);
\draw[->] (U) -- (D);
\draw[->] (U) -- (Y);
\draw[->] (V) -- (Z);
\draw[->] (V) -- (D);
\draw[->] (X) -- (D);
\draw[->] (X) -- (Y);

% nodes
\node (D) at (16,0) {$D$};
\node[circle,draw,dashed,inner sep=2pt] (U) at (17.5,-2.5) {$U$};
\node (Y) at (19,0) {$Y$};
\node[circle,draw,dashed,inner sep=2pt] (V) at (13,0) {$V$};
\node (Z) at (14.5,2.5) {$Z$};
\node (X) at (17.5,2.5) {$X$};

% arrows
\draw[->] (D) -- (Y);
\draw[->] (U) -- (D);
\draw[->] (U) -- (Y);
\draw[->] (V) -- (Z);
\draw[->] (V) -- (D);
\draw[->] (X) -- (D);
\draw[->] (X) -- (Y);
\end{tikzpicture}
\end{center}

```{r simulation DGP 1, message=FALSE}
library(AER)
library(ggplot2)
library(dplyr)
library(tidyr)
set.seed(1)

## run simulation for 200 times
results <-
  data.frame(iteration=1:200,
           OLS=rep(NA,200),
           IV=rep(NA,200))
for (i in 1:200){
  
  ## define unobserved
  V <- rnorm(1000, mean=3, sd=1)
  U <- rnorm(1000, mean=-2, sd=3)
  
  ## define observed
  Z <- 1 * V + runif(1000, min=0, max=2) + rnorm(1000, mean=0, sd=2)
  X <- rnorm(1000, mean=5, sd=3) + rnorm(1000, mean=0, sd=1)
  D <- -5 * V + 2 * Z + 0.5 * U + 1 * X + rnorm(1000, mean=0, sd=1)
  Y <- 5 * D + 1 * X + 2 * U + rnorm(1000, mean=0, sd=1)
  
  ## save results
  results[i,"OLS"] <- summary(lm(Y ~ D + X))$coef[2,1]
  results[i,"IV"] <- summary(ivreg(Y ~ X + D | X + Z))$coef[3,1]
}

## plot distribution
results %>%
  pivot_longer(c(OLS, IV), names_to = "Method", values_to = "Coefficient") %>%
  group_by(Method) %>%
  mutate(mean_coef = mean(Coefficient)) %>%
  ggplot(aes(x = Coefficient, fill = Method)) +
  geom_density(alpha = 0.4) +
  geom_vline(aes(xintercept = mean_coef, color = Method),
             linetype = "dashed", size = 1) +
  labs(title = "Distribution of Simulated Coefficients",
       x = "Coefficient", y = "Density") +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text()
  )
```

The simulation results show that you don't have to control for $V$ (it is unobserved in any case) or $X$ ($X$ does not affect $Z$) to use $Z$ as a valid IV, as long as $V$ does not affect $Y$.

Now what about another scenario where the relationship between $Z$ and $D$ is completely driven by unobserved $V$, as in the right panel? Yes you still can!

```{r simulation DGP 2, message=FALSE}
set.seed(1)

## run simulation for 200 times
results <-
  data.frame(iteration=1:200,
           OLS=rep(NA,200),
           IV=rep(NA,200))
for (i in 1:200){
  
  ## define unobserved
  V <- rnorm(1000, mean=3, sd=1)
  U <- rnorm(1000, mean=-2, sd=3)
  
  ## define observed
  Z <- 1 * V + runif(1000, min=0, max=2) + rnorm(1000, mean=0, sd=2)
  X <- rnorm(1000, mean=5, sd=3) + rnorm(1000, mean=0, sd=1)
  D <- -5 * V + 0.5 * U + 1 * X + rnorm(1000, mean=0, sd=1)
  Y <- 5 * D + 1 * X + 2 * U + rnorm(1000, mean=0, sd=1)
  
  ## save results
  results[i,"OLS"] <- summary(lm(Y ~ D + X))$coef[2,1]
  results[i,"IV"] <- summary(ivreg(Y ~ X + D | X + Z))$coef[3,1]
}

## plot distribution
results %>%
  pivot_longer(c(OLS, IV), names_to = "Method", values_to = "Coefficient") %>%
  group_by(Method) %>%
  mutate(mean_coef = mean(Coefficient)) %>%
  ggplot(aes(x = Coefficient, fill = Method)) +
  geom_density(alpha = 0.4) +
  geom_vline(aes(xintercept = mean_coef, color = Method),
             linetype = "dashed", size = 1) +
  labs(title = "Distribution of Simulated Coefficients",
       x = "Coefficient", y = "Density") +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5),
    axis.text.x = element_text()
  )
```