---
title: "\\textbf{Causal Machine Learning: Homework 2}"
subtitle: "\\textit{Due date}: October 6, 11:59PM"
author: "Your name:"
format:
  pdf:
    pdf-engine: pdflatex 
    documentclass: article
    classoption: ["12pt"]
    geometry: "top=1.1in, bottom=1.1in, left=1.1in, right=1.1in"
    include-in-header:
      text: |
        \DeclareMathOperator*{\argmin}{arg\,min}
        % fonts, math
        \usepackage[T1]{fontenc}
        \usepackage{dsfont}
        \usepackage{newpxtext}
        \let\arrowvert\relax
        \let\Arrowvert\relax
        \let\mathscr\relax
        \usepackage{newpxmath}
        \usepackage{amsmath}
        \usepackage{bm}
        
        % colors
        \usepackage[dvipsnames]{xcolor}
        
        % Indent & line spacing
        \usepackage{setspace}
        \setstretch{1.2}
        \setlength{\parindent}{20pt}
        \AtBeginDocument{
        \setlength{\abovedisplayshortskip}{-10pt}
        \setlength{\belowdisplayshortskip}{3.5pt}
        }
        
        % Figures
        \usepackage{graphicx}
        \usepackage{tikz}
        \usetikzlibrary{arrows.meta, positioning, fit}

        % Equal spacing between Title, Date, Author
        \makeatletter
        \renewcommand{\maketitle}{%
          \begin{center}
            {\LARGE \bfseries \@title \par}
            \vspace{0.75em}%
            {\large \itshape \@date \par}
            \vspace{0.75em}%
            {\large \@author \par}
          \end{center}\par\vspace{1em}
        }
        \makeatother
execute:
  echo: true
  warning: false
  message: false
editor: source
---

```{r load package, warning=FALSE, message=FALSE}
library(hdm)
library(keras3)
library(dplyr)
library(sandwich)
```

# Potential Outcome Framework

\noindent 1. In lecture, we showed that the observed difference in means (Average Predictive Effect, APE)

$$
\pi = E[Y_i \mid D_i=1] - E[Y_i \mid D_i=0]
$$

\noindent can be decomposed into the Average Treatment Effect on the Treated (ATT) plus a selection bias term. Show that $\pi$ can equivalently be decomposed into the Average Treatment Effect on the Controls (ATC) plus a selection bias term. In words, explain the meaning of this alternative selection bias.

\noindent 2. Suppose treatment status $D_i$ is independent of potential outcomes $Y_i(d), \:d\in\{0,1\}$ conditional on a set of covariates $X_i$. Define the Conditional ATE (CATE) as

$$
\delta(X_i) = E[Y_i(1) \mid X_i] - E[Y_i(0) \mid X_i]
$$

\noindent Show that

$$
\begin{aligned}
\delta(X_i) &= E\left[Y_i \left(\frac{\mathds{1}(D_i=1)}{P(D_i=1\mid X_i)} - \frac{\mathds{1}(D_i=0)}{1-P(D_i=1\mid X_i)}\right) \middle| X_i \right]
\end{aligned}
$$

\noindent where $P(D_i=1\mid X_i)$ is the probability of having $D_i=1$ given $X_i$.

# Directed Acyclic Graphs (DAG)

\noindent 1. The DAGs below graphically describe 3 hypothetical Structural Equation Models (SEMs). state which set of covariates need to be included to identify the causal effect of $B$ on $Y$.

\begin{center}
\begin{tikzpicture}[
  >=stealth,
  scale=0.65,
  thick,
  every node/.style={scale=1.1}]
  
  % DAG 1
  %% nodes
  \node (S) at (0,0) {$S$};
  \node (B) at (2,-2) {$B$};
  \node (Y) at (5,-2) {$Y$};
  \node (U) at (0,-4) {$U$};
  
  %% Arrows
  \draw[->] (S) -- (B);
  \draw[->] (S) -- (Y);
  \draw[->] (U) -- (B);
  \draw[->] (U) -- (Y);
  \draw[->] (B) -- (Y);
  \draw[->] (S) -- (U);
  \node at (2, -6) {\footnotesize Case 1};
  
  % DAG 2
  %% Nodes
  \node (S) at (8,0) {$S$};
  \node (B) at (10,-2) {$B$};
  \node (Y) at (13,-2) {$Y$};
  \node (U) at (8,-4) {$U$};
  
  %% Arrows
  \draw[->] (B) -- (U);
  \draw[->] (B) -- (S);
  \draw[->] (S) -- (Y);
  \draw[->] (S) -- (U);
  \draw[->] (U) -- (Y);
  \draw[->] (B) -- (Y);
  \node at (10, -6) {\footnotesize Case 2};
  
  % DAG 3
  %% Nodes
  \node (S) at (16,0) {$S$};
  \node (B) at (18,-2) {$B$};
  \node (Y) at (21,-2) {$Y$};
  \node (U) at (16,-4) {$U$};
  
  %% Arrows
  \draw[->] (S) -- (B);
  \draw[->] (S) -- (Y);
  \draw[->] (U) -- (B);
  \draw[->] (U) -- (Y);
  \draw[->] (B) -- (Y);
  \node at (18, -6) {\footnotesize Case 3};
\end{tikzpicture}
\end{center}

\noindent 2. Now focus on Case 3. Suppose the SEM has the following data-generating process

$$
\begin{aligned}
    Y &:= S + B + \kappa U + \epsilon_Y \\
    B &:= S + U + \epsilon_B \\
    S &:= \epsilon_S \\
    U &:= \epsilon_U
\end{aligned}
$$

\noindent where $\epsilon_U$, $\epsilon_Y$, $\epsilon_S$, and $\epsilon_B$ are independent $\mathcal{N}(0,1)$ shocks. Show the following equation holds, and explain the implication of the equation.

$$
E[Y \mid S, B] = (1-\kappa/2)S + (1+\kappa/2)B
$$

# Test of Convergence Hypothesis

In lecture, we provided an empirical example of partialling-out with Double LASSO to estimate the regression coefficient $\beta_1$ in the high-dimensional linear regression model. Specifically, we estimated how the rates at which economies of different countries grow ($Y_i$) are related to the initial wealth levels in each country ($D_i$) controlling for country's institutional, educational, and other similar characteristics ($X_i$). The relationship is captured by $\beta_1$, the *speed of convergence/divergence*, which measures the speed at which poor countries catch up $(\beta_1<0)$ or fall behind $(\beta_1>0)$ rich countries, after controlling for $X_i$.

In this assignment, I want you to use the formal procedure of Double Machine Learning to re-estimate $\beta_1$. Specifically, I want you fine-tune a shallow one-layer Neural Network to estimate the nuisance function for $Y_i$ and $D_i$ using $X_i$, and residualize $Y_i$ and $D_i$ using cross-fitting.

```{r load data, warning=FALSE}
library(hdm)
## function to get data
getdata <- function(...) {
  e <- new.env()
  name <- data(..., envir = e)[1]
  e[[name]]
}

## now load your data calling getdata()
growth <- getdata(GrowthData)
```

```{r create variables, warning=FALSE}
## create the outcome variable y, treatment d, and control covariates x
y <- growth$Outcome
x <- growth[-which(colnames(growth) %in% c("Outcome", "intercept", "gdpsh465"))]
d <- growth$gdpsh465
```

\noindent I have already implemented a Neural Network model using `keras3` and performed a manual grid search to fine-tune its hyperparameters. The tuning procedure evaluated each candidate model on a 20% validation split, and the best configuration was chosen based on the lowest validation loss. The default hyperparameter grid includes:

-   Number of hidden neurons (`units`): 10, 50, 100
-   Learning rate (`lr`): 0.001, 0.02, 0.05
-   Batch size: 15 (no tuning)
-   Number of epochs: 50, 100, 200, 400

This setup allowed the model to be optimized over different network sizes, learning speeds, mini-batch sizes, and training durations. I also used early stopping based on 10 epoch's validation loss.

```{r, fine tune NN, warning=FALSE}

## build and fit a keras model
build_and_fit <- function(x, y, units, lr, batch_size, epochs) {
  model <- keras_model_sequential() |>
    layer_dense(units = units, activation = "relu", input_shape = c(ncol(x))) |> ## hidden layer is controlled by units
    layer_dense(units = 1) |>
    compile(
      optimizer = optimizer_adam(learning_rate = lr), ## default optimizer is Adam
      loss = "mse"
    )
  
  ## fit the model
  history <- model %>% fit(
    x = as.matrix(x),
    y = y,
    batch_size = batch_size,
    epochs = epochs,
    validation_split = 0.2,
    callbacks = list(
      callback_early_stopping(
        monitor = "val_loss",
        patience = 10, ## early stopping criterion
        restore_best_weights = TRUE
      )
    ),
    verbose = 0
  )
  
  ## return the loss of the last validation set 
  val_loss <- tail(history$metrics$val_loss, 1)
  
  list(model = model, val_loss = val_loss)
}

## grid search for hyperparameters
tune_keras <- function(x, y, 
                       units_grid = c(10, 50, 100),
                       lr_grid = c(0.001, 0.02, 0.05),
                       batch_grid = c(15),
                       epoch_grid = c(50, 100, 200, 400)) {
  results <- expand.grid(units = units_grid,
                         lr = lr_grid,
                         batch_size = batch_grid,
                         epochs = epoch_grid)
  results$val_loss <- NA
  
  for (i in 1:nrow(results)) {
    fit <- build_and_fit(x, y,
                         units = results$units[i],
                         lr = results$lr[i],
                         batch_size = results$batch_size[i],
                         epochs = results$epochs[i])
    results$val_loss[i] <- fit$val_loss
  }
  
  ## find the best combination of hyperparameter
  best <- results[which.min(results$val_loss), ]
  return(best)
}
```

\noindent Now I want you to fine-tune the NN model *before* cross-fitting. That is, do not fine-tune the model within each $K-1$ fold, but only fine-tune model once based on the whole sample. Use the default grid search for $Y$, but use the following grid search for $D$:

-   Number of hidden neurons (`units`): 10, 50, 100
-   Learning rate (`lr`): 0.001, 0.02, 0.05
-   Batch size: 15 (no tuning)
-   Number of epochs: 500, 1000, 2000

Then use the NN model with the combination of hyperparameters that reports the lowest validation loss to estimate the nuisance function. You may find such model by calling

``` r
ymodel <- build_and_fit(
      x[-I[[b]], ], y[-I[[b]]], ## suppose you are retrieving the nuisance function for Y
      units = best_y$units,
      lr = best_y$lr,
      batch_size = best_y$batch_size,
      epochs = best_y$epochs
    )$model
```

Residualize your $Y$ and $D$ and report the point estimate of $\beta_1$ and its standard error using the standard DML pipeline. What can you conclude from this practice? Note this may take a while for `R` to complete.

```{r DML using NN, warning=FALSE, eval=FALSE}

## DML using NN - your code here
dmlNN_for_plm <- function(x, d, y, nfold = 5) {
  
}

## execute your DML NN function
set.seed(1)
dmlNN_for_plm(x, d, y, nfold = 5)
```
